---
title: "Module 2 — Intro to R and Quarto"
format: html
editor: source
execute:
  warning: false
  message: false
---

This is your workspace for the Module 2 worksheet. The full instructions live
on the lesson page — keep it open alongside this notebook. You will build up
this notebook one part at a time: fill in the empty chunk under each heading,
press **Run Cell** to see what happens, and write a sentence of your own
between the chunks. Render with **Preview** whenever the worksheet tells you to.

Always load the packages you need in a setup chunk at the top. Run this first,
every time you open the notebook:

```{r setup}
library(tidyverse)
library(here)
```

## Part 1 — Run a chunk, write a sentence, render


# R treats ** 'to the power off' 2 ** 10 is 2 multiplied by itself 10 times

# The brackets changed the order in which the code is being processed (BODMAS)
```{r}
3 + 5 + 10
10 * (5 + 1)
2 ** 10

64 / 2 + 2
64 / (2 + 2)
```

## Part 2 — Read the data and look at it

```{r}
table_02 <- read_csv(here("data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))

table_02

glimpse (table_02)
```

# There are 132 rows (each PIDs) and 6 columns (pid, timepoint, arm, nugent_score, crp_blood, ph)

## Part 3 — Pull out columns and summarise them

```{r}
table_02$nugent_score

mean(table_02$crp_blood, na.rm = TRUE) # skips missing values 
mean(table_02$crp_blood)

# removing na.rm = true did not change the value of the mean since there probably was not any missing values

crp_mean <- mean(table_02$crp_blood, na.rm = TRUE)
crp_mean

#the crp_mean 1.470379
```

## Part 4 — Build something with Pi

```{r}
# Load the data                                                                        
   library(dplyr)                                                                         
                                                                                          
   measurements <-                                                                        
 read.csv("data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv")                                                                                                                                                                     
   # Check it loaded okay                                                                 
   head(measurements)                                                                                                                                                   
   # Mean nugent_score by arm                                                             
   measurements |>                                                                        
     group_by(arm) |>                                                                     
     summarise(mean_nugent = mean(nugent_score, na.rm = TRUE)) 
```

## Capstone — make it your own

```{r}
# your code here
```

## Before you go

Pick one step today that confused you. Ask Pi to explain it, then push back:
*"why did you use X instead of Y?"* Notice when Pi sounds confident versus when
it is guessing.
