---
title: "Module 3 — Data wrangling"
format: html
editor: source
execute:
  warning: false
  message: false
---

> *Your mode for this exercise.* Group A: **agentic** — describe each step to Pi and let it write the code. Group B: **manual** — write each step yourself. A chat LLM may answer conceptual questions but should not edit this notebook or run commands. Exercise 2 flips the modes.

```{r setup}
library(tidyverse)
library(here)

table_02 <- read_csv(here("data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))
```

## Part 1 — Predict (no code)

Take a look at `table_02` (run the cell above first). Answer in plain words below — no code yet.

```{r}
table_02
```

**1. How many rows? What does one row represent?**

* 132 rows with each row representing a PID at one visit timepoint *

**2. After `table_02 |> filter(time_point == "baseline")`, how many rows? What does each row represent now?**

* 44 rows, one for each participant at their starting point (baseline) *

**3. After `table_02 |> count(arm)`, how many rows? What does each row represent now?**

* 2 rows with each row representing placebo which has 69 rows or treatment which has 63 rows *

## Part 2 — Build the pipeline

Each step adds one verb on top of the previous pipeline. Run after each step.

```{r}
table_02 |> head()
```

```{r}
table_02 |>
filter(time_point == "baseline")
```

```{r}
table_02 |>
select(pid, arm, ph, nugent_score, crp_blood)
```

```{r}
table_02 |> mutate(high_ph = if_else(ph >= 4.5, "yes", "no"))
```

```{r}
table_02 |> arrange(desc(ph))
```

```{r}
# step 6: assign the final pipeline to early_visit, then print it
early_visit <- table_02 |> filter(time_point == "baseline")
early_visit
```

```{r}
table_02 |> count(arm) 
```

## Part 3 — Reach (optional)

Add a `nugent_category` column with `case_when()`:

- 0–3 → "normal"
- 4–6 → "intermediate"
- 7–10 → "BV"

Then run `count(arm, nugent_category)`.

```{r}
# reach: mutate(nugent_category = case_when(...)) then count(arm, nugent_category)
```
