---
title: "Group A — Yogurt Study Data Dictionary"
subtitle: "Understanding every column in the dataset"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
editor: source
---

## Overview

This document describes every CSV file in the **Group A (yogurt study)** dataset.
There are four files, linked together by participant IDs (`pid`) and sample IDs
(`sample_id`).

```{r}
#| label: setup
#| message: false

library(tidyverse)

# Helper to read CSVs cleanly
read_data <- function(path) {
  read_csv(file.path("../../data/group_a_yogurt", path), show_col_types = FALSE)
}
```

---

## File 0 — Sample IDs (`00_sample_ids_yogurt.csv`)

```{r}
#| label: load-sample-ids

sample_ids <- read_data("00_sample_ids_yogurt.csv")
sample_ids
```

### What one row represents

**One biological sample** collected from a participant at a specific time point.
Each of the 51 participants contributes **two** samples (baseline + after
antibiotic), for a total of 102 rows.

### Columns

| Column       | Type   | What it means |
|--------------|--------|---------------|
| `pid`        | chr    | Participant ID — links to the metadata table (File 1) |
| `time_point` | chr    | When the sample was collected — either `baseline` or `after_antibiotic` |
| `arm`        | chr    | Which group the participant was assigned to — `yogurt` or `unchanged_diet` |
| `sample_id`  | chr    | Unique sample code — links to the qPCR (File 2) and Luminex (File 3) tables |

```{r}
#| label: explore-sample-ids

# How many samples per time point and arm?
sample_ids |>
  count(time_point, arm)
```

---

## File 1 — Participant Metadata (`01_participant_metadata_yogurt.csv`)

```{r}
#| label: load-metadata

metadata <- read_data("01_participant_metadata_yogurt.csv")
metadata
```

### What one row represents

**One participant** (a woman enrolled in the study). There are 51 participants.

### Columns

| Column               | Type   | What it means |
|----------------------|--------|---------------|
| `pid`                | chr    | Participant ID — links to the sample IDs table |
| `arm`                | chr    | Group assignment — `yogurt` or `unchanged_diet` (26 yogurt, 25 unchanged) |
| `days_since_last_sex`| num   | Number of days since the participant's most recent sexual activity (range: 1–48) |
| `birth_control`      | chr    | Type of birth control used — `no hormonal birth control` or `Depoprovera` |
| `age`                | num   | Age in years (range: 24–37) |
| `education`          | chr    | Highest education level — one of `less than grade 9`, `grade 10-12, matriculated`, `grade 10-12, not matriculated`, or `post-secondary` |
| `sex`                | chr    | All participants are `female` |

```{r}
#| label: explore-metadata

metadata |>
  count(arm, birth_control)

metadata |>
  summarise(
    n_participants = n(),
    age_min = min(age),
    age_max = max(age),
    age_mean = mean(age),
    days_since_sex_min = min(days_since_last_sex),
    days_since_sex_max = max(days_since_last_sex)
  )

metadata |>
  count(education)
```

---

## File 2 — qPCR Results (`02_qpcr_results_yogurt.csv`)

```{r}
#| label: load-qpcr

qpcr <- read_data("02_qpcr_results_yogurt.csv")
qpcr
```

### What one row represents

**Quantitative PCR measurements for one sample.** There are 102 rows — one per
sample.

### Columns

| Column           | Type   | What it means |
|------------------|--------|---------------|
| `sample_id`      | chr    | Sample code — links to the sample IDs table |
| `qpcr_bacteria`  | num   | Total bacterial load measured by qPCR (16S rRNA gene copies) |
| `qpcr_crispatus` | num   | Abundance of *Lactobacillus crispatus* (a beneficial vaginal bacterium) |
| `qpcr_iners`     | num   | Abundance of *Lactobacillus iners* (another *Lactobacillus* species common in the vaginal microbiome) |

```{r}
#| label: explore-qpcr

# Join with sample IDs to see which arm each qPCR result belongs to
qpcr_with_arm <- sample_ids |>
  select(sample_id, time_point, arm) |>
  inner_join(qpcr, by = "sample_id")

qpcr_with_arm |>
  group_by(time_point, arm) |>
  summarise(
    n = n(),
    mean_total_bacteria = mean(qpcr_bacteria),
    mean_crispatus = mean(qpcr_crispatus),
    mean_iners = mean(qpcr_iners),
    .groups = "drop"
  )
```

---

## File 3 — Luminex Cytokine Results (`03_luminex_results_yogurt.csv`)

```{r}
#| label: load-luminex

luminex <- read_data("03_luminex_results_yogurt.csv")
luminex
```

### What one row represents

**One cytokine measurement for one sample.** The table is in **long (tidy)
format**: each row is a single concentration value. There are 10 cytokines ×
102 samples = 1020 rows.

### Columns

| Column      | Type   | What it means |
|-------------|--------|---------------|
| `sample_id` | chr    | Sample code — links to the sample IDs table |
| `cytokine`  | chr    | Name of the cytokine or chemokine measured — one of: `IL-1a`, `IL-1b`, `IL-6`, `IL-8`, `IL-10`, `TNFa`, `IFN-Y`, `IP-10`, `MIG`, `MIP-3a` |
| `conc`      | num   | Concentration of that cytokine (units depend on the assay, e.g. pg/mL) |
| `limits`    | chr    | Whether the concentration falls within the detectable range — `within limits` or `out of range` |

```{r}
#| label: explore-luminex

# How many samples per cytokine, and how many are out of range?
luminex |>
  count(cytokine, limits) |>
  pivot_wider(names_from = limits, values_from = n)

# Join with arm info and summarise
luminex_with_arm <- sample_ids |>
  select(sample_id, time_point, arm) |>
  inner_join(luminex, by = "sample_id")

luminex_with_arm |>
  group_by(time_point, arm, cytokine) |>
  summarise(
    mean_conc = mean(conc, na.rm = TRUE),
    .groups = "drop"
  ) |>
  slice_max(mean_conc, n = 5)
```

---

## How the files connect

Here is a diagram of how the four tables relate to each other:

```
 ┌─────────────────────┐
 │ File 0: sample_ids  │ ◄── pid ────┐
 │  (102 rows, 4 cols) │             │
 └─────┬──────┬───────┘             │
       │      │                      │
  sample_id  sample_id               │
       │      │                      │
       ▼      ▼                      │
 ┌─────────┐ ┌──────────────┐  ┌────┴──────────────────┐
 │ File 2: │ │ File 3:      │  │ File 1: metadata       │
 │ qPCR    │ │ Luminex      │  │ (51 rows, 7 cols)      │
 │ (102 r) │ │ (1020 r)     │  │                         │
 └─────────┘ └──────────────┘  └─────────────────────────┘
```

- **File 1** holds one row per **participant** (51 participants).
- **File 0** links participants to their samples (2 samples each → 102 rows).
- **Files 2 and 3** hold lab measurements keyed by `sample_id`.

```{r}
#| label: join-all-tables
#| message: false
#| warning: false

# One big joined table for exploration
all_data <- sample_ids |>
  left_join(metadata, by = c("pid", "arm")) |>
  left_join(qpcr, by = "sample_id") |>
  left_join(luminex, by = "sample_id")

glimpse(all_data)
```

---

## Glossary of abbreviations

| Abbreviation | Full meaning |
|--------------|---------------|
| qPCR         | Quantitative polymerase chain reaction — a lab method to measure DNA amounts |
| *L. crispatus* | *Lactobacillus crispatus* — a bacterium linked to healthy vaginal microbiomes |
| *L. iners*     | *Lactobacillus iners* — another common vaginal *Lactobacillus* species |
| Luminex      | A bead-based assay that measures many proteins (like cytokines) at once |
| Cytokine     | A small signalling protein released by immune cells |
| Chemokine    | A type of cytokine that attracts immune cells to a site of infection |
| IL           | Interleukin (a family of cytokines) |
| TNFa         | Tumour necrosis factor alpha (a cytokine involved in inflammation) |
| IFN-Y        | Interferon gamma (a cytokine important for immune defence) |
| IP-10        | Interferon gamma-induced protein 10 (a chemokine) |
| MIG          | Monokine induced by gamma interferon (a chemokine) |
| MIP-3a       | Macrophage inflammatory protein 3 alpha (a chemokine) |
| Depoprovera  | A hormonal injectable contraceptive (medroxyprogesterone acetate) |

---

## Table 1 — Participant characteristics by study arm

A **Table 1** is the standard way to start a medical or epidemiology paper:
it shows the characteristics of participants side-by-side by treatment group,
so readers can see if the groups are well-balanced.

Here we use `gtsummary::tbl_summary()` to summarise the participant metadata
by arm (yogurt vs unchanged diet), then `add_p()` to test whether any
differences could be due to chance alone. A p-value < 0.05 suggests a
real difference between the groups.

```{r}
#| label: table-1
#| message: false

library(gtsummary)

metadata |>
  select(-pid, -sex) |>     # drop the ID column and the constant \`sex\` column
  tbl_summary(
    by = arm,               # summarise each column, split by study arm
    statistic = list(age ~ "{mean} ({sd})")  # age: mean (SD) instead of median (IQR)
  ) |>
  add_p(
    test = list(age ~ "t.test")              # age: t-test instead of Wilcoxon
  )
```

### What to look for

- **N (%)** rows for categorical variables like `birth_control` and `education`.
- **Mean (SD)** or **Median (IQR)** rows for continuous variables like `age` and
  `days_since_last_sex` (gtsummary automatically picks the right summary).
- The **p-value** column tells you whether the yogurt group and the
  unchanged-diet group differ significantly on that variable. If all p-values
  are large (> 0.05), the randomisation worked well.

```{r}
#| label: fig-birth-control-by-arm
#| message: false
#| fig.cap: "Birth control use in each study arm. p-value from a chi-squared test."

library(ggpubr)

# Chi-squared test on the arm × birth_control table
chisq <- chisq.test(table(metadata$arm, metadata$birth_control))

# Build the bar plot — counts of each birth control type, side-by-side by arm
metadata |>
  count(arm, birth_control) |>
  ggbarplot(
    x = "arm",
    y = "n",
    fill = "birth_control",
    position = position_dodge(0.8),
    xlab = "Study arm",
    ylab = "Number of participants"
  ) +
  annotate("text",
    x = 1.5, y = max(table(metadata$arm, metadata$birth_control) + 1),
    label = paste0("chi-squared p = ", round(chisq$p.value, 3)),
    size = 4
  )
```

Because `birth_control` is a **categorical** variable, a Wilcoxon test doesn't
apply here. Instead we use a **chi-squared test**, which checks whether the
proportion of Depoprovera users is the same in the yogurt and unchanged-diet
arms.

---

## Research question — Effect of yogurt on the vaginal microbiome after antibiotics

> **Question:** Does eating yogurt (vs. an unchanged diet) help restore a
> healthier vaginal microbiome after antibiotic treatment?

To answer this, we look at qPCR measurements of three things:

- **Total bacteria** — overall bacterial load
- ***L. crispatus*** — a bacterium linked to a healthy vaginal microbiome
- ***L. iners*** — another common *Lactobacillus* species

We compare **yogurt** vs. **unchanged diet** at each time point (baseline and
after antibiotic). The plots below show abundance on a log scale (adding 1 to
handle zeros). A **Wilcoxon rank-sum test** is used to compare the two groups
at each time point — non-significant p-values (ns) suggest that any visible
difference could just be due to chance.

```{r}
#| label: fig-qpcr-facet
#| message: false
#| fig.width: 12
#| fig.height: 6
#| fig.cap: "qPCR results faceted by bacterium. Boxplots compare yogurt vs unchanged diet at baseline and after antibiotic treatment. Pink = unchanged diet, purple = yogurt. p-values from Wilcoxon rank-sum test."

library(scales)

# Join qPCR with sample IDs, reshape to long format for faceting
qpcr_long <- sample_ids |>
  inner_join(qpcr, by = "sample_id") |>
  pivot_longer(
    cols = c(qpcr_bacteria, qpcr_crispatus, qpcr_iners),
    names_to = "bacterium",
    values_to = "abundance"
  ) |>
  mutate(
    bacterium = recode(bacterium,
      qpcr_bacteria = "Total bacteria",
      qpcr_crispatus = "L. crispatus",
      qpcr_iners = "L. iners"
    ),
    time_point = fct_relevel(time_point, "baseline", "after_antibiotic"),
    abundance = abundance + 1
  )

# Compute Wilcoxon p-values for each bacterium × time point
pvals <- qpcr_long |>
  group_by(bacterium, time_point) |>
  summarise(
    p = wilcox.test(abundance ~ arm, data = pick(everything()))$p.value,
    .groups = "drop"
  ) |>
  mutate(
    star = case_when(
      p < 0.001 ~ "***",
      p < 0.01  ~ "**",
      p < 0.05  ~ "*",
      TRUE      ~ "ns"
    ),
    label = paste0("p = ", format(p, digits = 3), " ", star)
  )

# Custom colours: unchanged diet = pink, yogurt = purple
arm_colors <- c("unchanged_diet" = "#D81B60", "yogurt" = "#7B1FA2")

ggplot(qpcr_long, aes(x = time_point, y = abundance)) +
  geom_boxplot(aes(fill = arm), alpha = 0.5, outlier.shape = NA,
               width = 0.5, position = position_dodge(width = 0.75)) +
  geom_point(aes(color = arm),
             position = position_dodge(width = 0.75),
             size = 1.2, alpha = 0.6) +
  facet_wrap(~ bacterium, scales = "free_y") +
  scale_fill_manual(values = arm_colors,
                    labels = c("Unchanged diet", "Yogurt")) +
  scale_color_manual(values = arm_colors, guide = "none") +
  scale_y_log10(labels = comma) +
  labs(x = "Time point",
       y = "Abundance (16S rRNA gene copies + 1)",
       fill = "Study arm") +
  geom_text(
    data = pvals,
    aes(x = time_point, y = Inf, label = label),
    vjust = 1.5, size = 3.2, inherit.aes = FALSE
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "grey95"),
    strip.text = element_text(face = "italic")
  )
```

### What to look for

- **Total bacteria:** Are the two arms similar at baseline? Does one group show
  a bigger drop after antibiotics?
- ***L. crispatus*:** This "good" bacterium often drops after antibiotics.
  Does it recover more in the yogurt group?
- ***L. iners*:** This species can be protective or harmful depending on the
  context. Does yogurt affect it differently than the unchanged diet?

The p-values above each pair of boxes tell you whether the difference between
yogurt and unchanged diet is statistically significant at that time point.

### Export

You can download standalone copies of the table and the figure from the files
below. The code chunk saves them using `gtsave()` (for the gtsummary table)
and `ggsave()` (for the ggplot / ggpubr figure).

```{r}
#| label: export-files
#| message: false

# Create the output folder if it doesn't exist already
dir.create("output", showWarnings = FALSE)

# --- Table 1 as HTML ---

library(gtsummary)
library(gt)  # for gtsave()

t1 <- metadata |>
  select(-pid, -sex) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})")
  ) |>
  add_p(
    test = list(age ~ "t.test")
  )

gtsave(as_gt(t1), "output/table-1.html")

# --- Figure 1 as PDF ---

library(ggpubr)

chisq <- chisq.test(table(metadata$arm, metadata$birth_control))

p <- metadata |>
  count(arm, birth_control) |>
  ggbarplot(
    x = "arm",
    y = "n",
    fill = "birth_control",
    position = position_dodge(0.8),
    xlab = "Study arm",
    ylab = "Number of participants"
  ) +
  annotate("text",
    x = 1.5, y = max(table(metadata$arm, metadata$birth_control) + 1),
    label = paste0("chi-squared p = ", round(chisq$p.value, 3)),
    size = 4
  )

ggsave("output/figure-1.pdf", plot = p, width = 6, height = 4)

# --- Figure 2 — qPCR facet plot as PDF ---

library(scales)

qpcr_long <- sample_ids |>
  inner_join(qpcr, by = "sample_id") |>
  pivot_longer(
    cols = c(qpcr_bacteria, qpcr_crispatus, qpcr_iners),
    names_to = "bacterium",
    values_to = "abundance"
  ) |>
  mutate(
    bacterium = recode(bacterium,
      qpcr_bacteria = "Total bacteria",
      qpcr_crispatus = "L. crispatus",
      qpcr_iners = "L. iners"
    ),
    time_point = fct_relevel(time_point, "baseline", "after_antibiotic"),
    abundance = abundance + 1
  )

pvals <- qpcr_long |>
  group_by(bacterium, time_point) |>
  summarise(
    p = wilcox.test(abundance ~ arm, data = pick(everything()))$p.value,
    .groups = "drop"
  ) |>
  mutate(
    star = case_when(
      p < 0.001 ~ "***",
      p < 0.01  ~ "**",
      p < 0.05  ~ "*",
      TRUE      ~ "ns"
    ),
    label = paste0("p = ", format(p, digits = 3), " ", star)
  )

arm_colors <- c("unchanged_diet" = "#D81B60", "yogurt" = "#7B1FA2")

p2 <- ggplot(qpcr_long, aes(x = time_point, y = abundance)) +
  geom_boxplot(aes(fill = arm), alpha = 0.5, outlier.shape = NA,
               width = 0.5, position = position_dodge(width = 0.75)) +
  geom_point(aes(color = arm),
             position = position_dodge(width = 0.75),
             size = 1.2, alpha = 0.6) +
  facet_wrap(~ bacterium, scales = "free_y") +
  scale_fill_manual(values = arm_colors,
                    labels = c("Unchanged diet", "Yogurt")) +
  scale_color_manual(values = arm_colors, guide = "none") +
  scale_y_log10(labels = comma) +
  labs(x = "Time point",
       y = "Abundance (16S rRNA gene copies + 1)",
       fill = "Study arm") +
  geom_text(
    data = pvals,
    aes(x = time_point, y = Inf, label = label),
    vjust = 1.5, size = 3.2, inherit.aes = FALSE
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "grey95"),
    strip.text = element_text(face = "italic")
  )

ggsave("output/figure-2.pdf", plot = p2, width = 12, height = 6)
```

| File | Description |
|------|-------------|
| [`output/table-1.html`](output/table-1.html) | Table 1 — participant characteristics by study arm (open in your browser) |
| [`output/figure-1.pdf`](output/figure-1.pdf) | Figure 1 — birth control use by study arm |
| [`output/figure-2.pdf`](output/figure-2.pdf) | Figure 2 — qPCR results: effect of yogurt on the vaginal microbiome after antibiotics |

---

## Linear model — adjusting for baseline

So far, we compared yogurt vs unchanged diet **separately** at baseline and
after antibiotics (the boxplots with Wilcoxon tests above). But that approach
doesn't use all the information we have — each participant was measured twice,
and knowing their baseline level helps us make a more precise comparison after
antibiotics.

A **linear model** (`lm()`) lets us ask a more targeted question:

> *After antibiotics, is the yogurt group different from the unchanged diet
> group — **adjusting for where each person started at baseline**?*

This is better because:
1. It **accounts for individual differences** — if one person naturally has
   higher bacterial levels, the model knows that from their baseline.
2. **One model, one answer** — a single coefficient tells you the yogurt effect.
3. **More statistical power** — using baseline as information makes it easier
   to detect a real effect.

### Step 1 — Widen the data

First we reshape the data so each participant has **one row** with separate
columns for their baseline and after-antibiotic measurements.

```{r}
#| label: model-widen-data
#| message: false

# Join qPCR with sample IDs and log-transform
qpcr_with_arm <- sample_ids |>
  inner_join(qpcr, by = "sample_id") |>
  mutate(
    log_bacteria  = log10(qpcr_bacteria + 1),
    log_crispatus = log10(qpcr_crispatus + 1),
    log_iners     = log10(qpcr_iners + 1)
  )

# Widen: one row per participant, baseline and after_antibiotic as columns
wide <- qpcr_with_arm |>
  select(pid, time_point, arm, log_bacteria, log_crispatus, log_iners) |>
  pivot_wider(
    names_from = time_point,
    values_from = c(log_bacteria, log_crispatus, log_iners)
  )

wide
```

### Step 2 — Fit the models

We fit one model for each bacterial outcome. The formula says:

> `after_antibiotic ~ arm + baseline`
>
> "Predict the after-antibiotic level from the study arm, **controlling for**
> the baseline level."

```{r}
#| label: model-fit
#| message: false

library(gtsummary)

# --- Total bacteria ---
m_total <- lm(log_bacteria_after_antibiotic ~ arm + log_bacteria_baseline,
              data = wide)

# --- L. crispatus ---
m_crispatus <- lm(log_crispatus_after_antibiotic ~ arm + log_crispatus_baseline,
                  data = wide)

# --- L. iners ---
m_iners <- lm(log_iners_after_antibiotic ~ arm + log_iners_baseline,
              data = wide)
```

### Step 3 — Turn into coefficient tables

`tbl_regression()` takes the fitted model and makes a clean table with the
coefficient (Beta), 95% confidence interval, and p-value for each predictor.

```{r}
#| label: model-tables
#| message: false

# --- Total bacteria ---
tbl_regression(m_total) |>
  modify_caption("**Model 1: Total bacteria (after antibiotics)**")

# --- L. crispatus ---
tbl_regression(m_crispatus) |>
  modify_caption("**Model 2: L. crispatus (after antibiotics)**")

# --- L. iners ---
tbl_regression(m_iners) |>
  modify_caption("**Model 3: L. iners (after antibiotics)**")
```

### How to read the tables

Each table has four columns:

| Column | What it tells you |
|---|---|
| **Characteristic** | The predictor variable in the model |
| **Beta** | The estimated effect — for `arm: yogurt`, this is the difference between the yogurt group and the unchanged diet group **after antibiotics, adjusting for baseline** |
| **95% CI** | The range of plausible values for the true effect. If it crosses 0, we can't be sure the effect isn't zero. |
| **p-value** | How surprised we'd be if there's truly no effect. Small (< 0.05) = statistically significant. Large (> 0.05) = could be due to chance. |

### What the results tell us

| Outcome | Yogurt effect (Beta) | 95% CI | p-value | Plain English |
|---|---|---|---|---|
| **Total bacteria** | −0.37 | −0.77 to 0.02 | **0.061** | Yogurt group had *slightly lower* total bacteria after antibiotics — **borderline** not significant |
| ***L. crispatus*** | +0.82 | −0.40 to 2.0 | **0.2** | Yogurt group had *somewhat higher* crispatus, but the wide confidence interval means we're very uncertain |
| ***L. iners*** | −0.17 | −0.78 to 0.44 | **0.6** | Basically no difference between groups |

**Key takeaway:** None of the effects reached statistical significance at
p < 0.05. But the total bacteria result (p = 0.061) is the most interesting
— it *hints* that yogurt might help reduce bacterial load after antibiotics.
Adjusting for baseline made this pattern clearer than the simple Wilcoxon
comparisons in Figure 2.

### Experiment to try

Try adding `age` or `birth_control` as an extra covariate to see if the
results change:

```r
# Example: adjust for age as well
lm(log_bacteria_after_antibiotic ~ arm + log_bacteria_baseline + age,
   data = wide) |>
  tbl_regression()
```

You'd need to join `metadata` into the `wide` table first (they share the `pid`
column). If the yogurt coefficient barely moves when you add age, that tells
you the randomisation worked — age is balanced between the two arms.